---
title: "Data Preparation Pipeline"
author: "Erik Westlund"
date: "2025-01-13"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
    html_document:
      keep_md: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(DT)
library(kableExtra)
library(purrr)
library(readr)
library(readxl)
library(rlang)
library(stringr)
library(tidyr)

```

```{r ingest_data, echo=FALSE, message=FALSE}

raw_analysis_data <- readRDS("data/FSCI_2024.rds")
raw_target_data <- readxl::read_xlsx("data/Codebook Dec 2024.xlsx", sheet = "Indicators and Target Value")
target_year <- 2030

# Contains functions for data preparation
source("functions.R")
source("data_patches.R")

# Stored in data_patches.R
data_patches <- get_data_patches()

```

# Countries

```{r country_summary, echo=FALSE, message=FALSE}
country_indicator_summary <- get_country_indicator_summary(raw_analysis_data)
```

## Summary

```{r country_counts, echo=FALSE, message=FALSE}
country_indicator_summary |> 
  summarize(
    n_countries = n(),
    n_regions = n_distinct(region),
    n_income_groups = n_distinct(income_group),
  ) |>
  kable()

```

## Indicators Per Year

The below table shows the number of non-missing indicators a country has in each year.
<br><br>
```{r country_indicators, echo=FALSE, message=FALSE}
DT::datatable(country_indicator_summary, rownames = FALSE, options = list(pageLength = 20, scrollX = TRUE))
```

# Regions

## Summary

```{r region_summary, echo=FALSE, message=FALSE}
region_summary <- get_region_summary(raw_analysis_data)

region_summary |> kable()
```

# Income Groups

## Summary

```{r income_group_summary, echo=FALSE, message=FALSE}
income_group_summary <- get_income_group_summary(raw_analysis_data)
income_group_summary |> kable()
```

# Indicators

```{r indicator_summary, echo=FALSE, message=FALSE}
indicator_summary <- get_indicator_summary(raw_analysis_data)
```

## Summary

```{r indicator_summary_table, echo=FALSE, message=FALSE}

indicator_summary |> 
  summarize(
    n_indicators = n(),
    first_year = min(first_year),
    latest_year = max(latest_year),
    total_years = latest_year - first_year + 1
  ) |>
  kable()

```

## Indicators 

```{r indicator_table, echo=FALSE, message=FALSE}

DT::datatable(indicator_summary, rownames = FALSE, options = list(pageLength = 60, scrollX = TRUE))

```

# Metrics

```{r calculate_metrics, echo=FALSE, message=FALSE, results='asis'}

metrics_data <- get_metrics_data(raw_analysis_data, indicator_summary, raw_target_data, target_year, data_patches)
# metrics_data <- readRDS("output/metrics_data.rds")

# Debug:
# indicator_variable_metrics <- metrics_data[[1]]
for(indicator_variable_metrics in metrics_data) {
  cat(paste0("## ", indicator_variable_metrics$variable, ": ", indicator_variable_metrics$short_label , "\n\n"))
  
  cat("\n\n")
  table <- summarize_metrics_collection(indicator_variable_metrics) |> kable()
  print(table) # Ensure the table is rendered in the output
  
}

```

# Save Output

Two files have been saved to the `output` folder:

* `metrics_data.rds` contains a list of lists, where each list contains a dataframe of the metrics for each indicator as well as meta-data.
* `metrics_combined_dataframe.rds` contains a data frame of all the metrics for each indicator combined into a single data frame.

```{r save_output, echo=FALSE}       

combined_data <- map_dfr(
  metrics_data,
  ~ .x$data,
  .id = "variable"
)

saveRDS(metrics_data, "output/metrics_data.rds")
saveRDS(combined_data, "output/metrics_combined_dataframe.rds")

```
