---
title: "Data Preparation Pipeline"
author: "Erik Westlund"
date: "2025-01-13"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
    html_document:
      keep_md: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(DT)
library(kableExtra)
library(purrr)
library(readr)
library(readxl)
library(rlang)
library(stringr)
library(tidyr)

```

```{r ingest_data, echo=FALSE, message=FALSE}

raw_data <- readRDS("data/FSCI_2024.rds")
targets <- readxl::read_xlsx("data/Codebook Dec 2024.xlsx", sheet = "Indicators and Target Value")
target_year <- 2030

# Contains maps for renaming variables and indicators for this pipeline.
# source("naming_maps.R")

# Contains functions for data preparation
source("functions.R")

```

# Data Preparation

Our focus for now is the value variable, so for each variable, we need to check for mixed types in the value column.  The below tables show the first five values for each variable.

We find that `foodenviropolicies` and `righttofood` are the only char vars.

## Data Type Analysis Diagnostics

```{r variable_cleanup, echo=FALSE, results='asis'}
data <- raw_data

diagnostics <- data |> 
  group_by(variable) |> 
  summarize(
    classification = list(classify_values_individually(value)),
    .groups = "drop"
  ) |> 
  mutate(
    ex1_num = map_chr(classification, ~ as.character(.x$numeric_examples[1] %||% NA)),
    ex2_num = map_chr(classification, ~ as.character(.x$numeric_examples[2] %||% NA)),
    ex3_num = map_chr(classification, ~ as.character(.x$numeric_examples[3] %||% NA)),
    ex4_num = map_chr(classification, ~ as.character(.x$numeric_examples[4] %||% NA)),
    ex5_num = map_chr(classification, ~ as.character(.x$numeric_examples[5] %||% NA)),
    ex1_char = map_chr(classification, ~ as.character(.x$char_examples[1] %||% NA)),
    ex2_char = map_chr(classification, ~ as.character(.x$char_examples[2] %||% NA)),
    ex3_char = map_chr(classification, ~ as.character(.x$char_examples[3] %||% NA)),
    ex4_char = map_chr(classification, ~ as.character(.x$char_examples[4] %||% NA)),
    ex5_char = map_chr(classification, ~ as.character(.x$char_examples[5] %||% NA)),
    full_numeric = map(classification, "numeric_full"),
    full_char = map(classification, "char_full")
  ) |> 
  select(variable, ex1_num:ex5_char, full_numeric, full_char)

cat("## Numeric Indicators\n\n")
cat("### Indicators with non-missing numeric data\n\n")

# Note: just check value 1; if it's NA, all are NA
num_vars <- diagnostics |> 
  filter(!is.na(ex1_num)) |>
  pull(variable)

num_vars |> kable()

cat("### Samples of values\n\n")

diagnostics |>
  select(variable, ex1_num:ex5_num) |> kable()

cat("## Character Indicators\n\n")
cat("### Indicators with non-missing character data\n\n")

char_vars <- diagnostics |> 
  filter( !is.na(ex1_char)) |>
  pull(variable)

char_vars |> kable()

cat("### Sample of Values\n\n")

diagnostics |>
  select(variable, ex1_char:ex5_char) |> kable()

cat("## Logical Variables\n\n")

cat("### Indicators with non-missing logical data\n\n")

logical_vars <- diagnostics |> 
  filter(ex1_num %in% c(0,1)) |>
  filter(ex2_num %in% c(0,1)) |>
  filter(is.na(ex3_num)) |> 
  pull(variable)

logical_vars |> kable()

cat("### Sample of Values\n\n")

diagnostics |>
  filter(variable %in% logical_vars) |>
  select(variable, ex1_num:ex5_num) |> kable()

cat("## Mixed Variables\n\n")

mixed_vars <- diagnostics |> 
  filter(!is.na(ex1_num) & !is.na(ex1_char)) |>
  pull(variable)

mixed_vars |> kable()

cat("### Sample of Values\n\n")

diagnostics |>
  filter(variable %in% mixed_vars) |>
  select(variable, ex1_num:ex5_num, ex1_char:ex5_char) |> kable()

## Char only vars
char_only_vars <- char_vars |> setdiff(mixed_vars)

```

### Notes

`croplandchange` is mixed; has an empty string, but mostly numeric values. This should not affect results since numeric operations will coerce this to NA.

## Dataframe Preparation

Create a dataframe named `analysis_long_data` with:

* Renamed columns for ease of debugging (lowercase, snake case between words, units/metrics as suffixes)
* Renamed `variable` column for ease of debugging (same rules as above)
* Non-numeric variables removed (i.e., `righttofood`, `foodenviropolicies`)

Note: upon finishing this process, renaming can be removed with little work. Alternatively, renaming can be kept and reversed. The results would be the same.

```{r data_cleaning, echo=FALSE}

analysis_long_data <- data |> 
  filter(!variable %in% c("righttofood", "foodenviropolicies")) |> 
  mutate(
    variable = clean_analysis_data_variables(variable),
  ) |> 
  mutate( # Utility variables to make it easier to work with columns with spaces
    fsci_region = FSCI_region,
    un_contentinal_region = `UN Continental Region`,
    milestone_pctile = if_else(desirable_direction == -1, 0.2, 0.8)
  )

regions <- get_regions(analysis_long_data)
income_groups <- get_income_groups(analysis_long_data)

target_data <- targets |> 
  rename(
    variable = `Variable name`,
    indicator = `Indicator`,
    unit = `Unit`,
    thematic_grouping = `Thematic Grouping`,
    has_target = `Target (yes/no)`,
    target = `Target Global (value)`,
  ) |> 
  mutate(
    has_target = if_else(has_target == "Yes", TRUE, FALSE)
  )

```



# Process Data & Calculate Metrics

```{r extract_data_and_milestones, echo=FALSE, results='asis'}

# Initialize a list to store data for each variable
metrics_data <- tibble()
failing_variables <- c()
warning_variables <- c()

indicator_variables <- get_indicator_variables(analysis_long_data, exclude = c("righttofood", "foodenviropolicies"))

for(indicator_variable in indicator_variables) {
  
  # Initialize in the event of an error
  indicator_metrics <- tibble()
  
  short_label <- target_data |> 
    filter(variable == indicator_variable) |> 
    pull(indicator)
  
  cat(paste0("Processing `", indicator_variable ,"`: ", short_label, " ...\n"))
  tryCatch(
    {
      indicator_data <- extract_indicator_data(analysis_long_data, indicator_variable)
      target_value <- get_target_value(indicator_variable, target_data)
      indicator_milestone_metrics <- calculate_milestone_metrics(indicator_data) # All years for milestones
      indicator_distance_metrics <- calculate_distance_metrics(indicator_data, indicator_milestone_metrics, target_value)
      indicator_velocity_metrics <- calculate_velocity_metrics(indicator_data, indicator_milestone_metrics, target_value, target_year)
      
      indicator_metrics <- indicator_milestone_metrics |> 
        mutate(
          target_value = target_value
        ) |> 
        inner_join(indicator_distance_metrics, by = c("variable", "country", c("year" = "latest_year"))) |>  # Narrows to 2022 
        rename(
          latest_year = year,
          latest_year_value = value,
        ) |> 
        select(variable, country, latest_year,  latest_year_value, everything()) |> # Rearrange
        inner_join(indicator_velocity_metrics |> select(-milestone_global, -milestone_region, -milestone_income_group), by = c("variable", "country", "latest_year")) 
      
      cat(paste0("âœ… Sucessfully processed.\n\n"))
    },
    warning = function(w) {
      warning_variables <- c(warning_variables, indicator_variable)
      cat(paste0("âš ï¸ Warning processing ", indicator_variable, ": ", w$message, "\n\n"))
    },
    error = function(e) {
      failing_variables <- c(failing_variables, indicator_variable)
      cat(paste0("ðŸ’£ Error processing ", indicator_variable, ": ", e$message, "\n\n"))
    }
  )
  
  metrics_data <- bind_rows(metrics_data, indicator_metrics)
}
```

# Milestone, Distance, and Velocity Metrics

```{r milestones_and_growth_rates, results='asis', echo=FALSE}

  DT::datatable(metrics_data, 
    options = list(
      pageLength = 30,
      scrollX = TRUE
    )
  )

readr::write_csv(metrics_data, "data/metrics_data.csv")



```