---
title: "Milestone Doublechecking"
author: "FSCI Group"
date: "2025-02-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(dplyr)
library(forcats)
library(ggplot2)
library(ggtext)
library(purrr)
library(readr)
library(readxl)
library(stringr)
library(tidyr)

library(rlang)
library(kableExtra)
library(DT)


# Contains functions for data preparation
source("functions.R")
source("data_patches.R")

# Stored in data_patches.R
data_patches <- get_data_patches()


```


# Load Data

```{r load_data}

raw_analysis_data <- readRDS("data/FSCI_2024.rds")
raw_target_data <- readxl::read_xlsx("data/Codebook Dec 2024.xlsx", sheet = "Indicators and Target Value")
target_year <- 2030

analysis_data <- get_analysis_data(raw_analysis_data) 

analysis_data |> head()

target_data <- get_target_data(raw_target_data)

target_data |> head()

indicator_variables <- get_indicator_variables(analysis_data)
indicator_variables |> print(print_n=Inf)

```

# Example 1: Global, Safe Water (safeh20)


## Direction

It's positive:

```{r example_1_global_safe_water_dir}

var <- 'safeh20'

desirable_direction <- analysis_data |> 
  filter(variable == var) |> 
  pull(desirable_direction) |> 
  unique()

desirable_direction

percentile <- ifelse(desirable_direction == 1, 80, 20)

percentile

```

## Data

```{r example_1_global_safe_water_data}

safeh20_global_data <- analysis_data |> 
  filter(variable == 'safeh20') |> 
  select(country, year, value) |> 
  arrange(country, year) |> 
  mutate(
    value = as.numeric(value)
  )

DT::datatable(safeh20_global_data, rownames = FALSE)


```

## Milestone calculation

We take the latest year with date.

```{r example_1_global_safe_water_milestone}

last_year_milestone_data <- safeh20_global_data |> 
  filter(!is.na(value)) |>
  group_by(country) |> 
  mutate(
    max_year = max(year),
  ) |> 
  filter(year == max_year) |>
  ungroup()

last_year_milestone_data |> kable()

```

Then we take all values over/under this percentile, depending on the direction.

```{r example_1_global_safe_water_milestone_2}

gt_lt <- ifelse(desirable_direction == 1, '>=', '<=')

global_milestone <- if(desirable_direction == 1) {
  last_year_milestone_data |> 
    filter(value >= quantile(value, 0.80)) |> 
    summarise(
      global_milestone = mean(value, na.rm = TRUE)
    ) 
} else {
  last_year_milestone_data |> 
    filter(value <= quantile(value, 0.20)) |> 
    summarise(
      global_milestone = mean(value, na.rm = TRUE)
    ) 
}

global_milestone |> kable()


```

Using the method of:

1. Take the latest year.
2. Take the 80th percentile if the direction is positive, or the 20th percentile if the direction is negative.
3. Calculate the mean of the values that are greater than or equal to the 80th percentile, or less than or equal to the 20th percentile.

We get a global milestone of `99.60`. This is still different from the provided Excel file, which has `98.77`. 


# Example 2: Region, Safe Water

## Direction

Same as above; leave alone.

## Data

```{r example_2_region_safe_water_data}

safeh20_region_data <- analysis_data |> 
  filter(variable == 'safeh20') |> 
  select(un_continental_region, country, year, value) |> 
  arrange(un_continental_region, country, year) |> 
  mutate(
    value = as.numeric(value)
  )

DT::datatable(safeh20_region_data, rownames = FALSE)

```

## Milestone calculation

```{r example_2_region_safe_water_milestone}

last_year_region_milestone_data <- analysis_data |> 
  filter(variable == 'safeh20') |> 
  select(country, year, value, un_continental_region) |> 
  mutate(
    value = as.numeric(value)
  ) |> 
  filter(!is.na(value)) |>
  group_by(country) |> 
  mutate(
    max_year = max(year),
  ) |> 
  filter(year == max_year) |>
  ungroup()

region_milestone_data <- if(desirable_direction == 1) {
  last_year_region_milestone_data |> 
    group_by(un_continental_region) |>
    filter(value >= quantile(value, 0.80)) |> 
    summarise(
      value = mean(value, na.rm = TRUE)
    ) 
} else {
  last_year_milestone_data |> 
    group_by(un_continental_region) |>
    filter(value <= quantile(value, 0.20)) |> 
    summarise(
      value = mean(value, na.rm = TRUE)
    )
}

region_milestone_data |> kable()

```

The above values differ from what is calculated in the Excel files, where:

* Africa = `54.5878295898437`
* Americas = `94.588737487793`
* Asia = `97.8738479614258`
* Europe = `99.8041534423828`
* Oceania = `99.5280838012695`

